---
title: "9 - exercise"
format: 
  html: 
    embed-resources: true
editor: visual
---

Pick a ticker - any ticker. For this exercise you'll compute the full performance dashboard and run a simple moving-average crossover backtest against buy-and-hold. At the end, in plain English, tell me which you'd prefer to hold and why.

### 1. Load libraries

```{r}
#| label: load-libraries
#| warning: false
#| message: false


```

### 2. Grab 15 years of daily data and compute returns

```{r}
#| label: grab-data
#| warning: false


```

### 3. Compute the core metrics

Write a function `perf_summary(r, rf_annual = 0.04)` that takes a vector of daily returns and returns a one-row tibble with annualized return, annualized volatility, Sharpe, Sortino, and max drawdown. Apply it to your buy-and-hold return series.

```{r}
#| label: perf-summary-fn


```

### 4. Drawdown chart

Plot the drawdown (equity curve divided by its running max, minus 1) for buy-and-hold. Shade it red.

```{r}
#| label: drawdown-chart


```

### 5. Backtest a 50/200 SMA crossover

Build signal and strategy returns. Remember to lag the signal by one day to avoid look-ahead bias.

```{r}
#| label: backtest


```

### 6. Head-to-head table

Stack buy-and-hold and the strategy in a single tibble using your `perf_summary` function so you can compare side by side.

```{r}
#| label: head-to-head


```

### 7. Equity curves

Plot both equity curves on a log scale so both sides of the chart are readable.

```{r}
#| label: equity-curves


```

### 8. Interpretation

Two to four sentences. Which would you rather hold? Why? What are the limits of what this test actually demonstrates?

**ANSWER:**
